Something fancy...

### 2020-01-27 draft.tex
Update inhomogeneous

\section{Introduction}
Topological phases of matter have attracted more and more attention during the last decades, not the least because many more relevant systems have been realized experimentally. 
The early focus was mainly on \emph{topologically ordered} systems \cite{Wen}, where interaction effects are crucial for stabilizing the phase. 
The most notable examples are the fractional quantum Hall effect~\cite{} and quantum spin liquids~\cite{}. 
However, since 2005 \maria{?}~\cite{KaneMele, etc} the focus has shifted to non-interacting topological phases, which can be characterized in terms of topological invariants~\cite{}. 
Symmetries are often necessary to protect these topological phases, and determine which distinct topological phases can be realized for a given dimensionality.

XXX
How to identify interesting phases?--- entanglement entropy, entanglement spectrum

Kitaev honeycomb model as example of entanglement spectrum

What information is encoded in the ES? How to extract it?

naively resp. easily accessible: information about edge spectrum

But, other information can be accessed as well : e.g. compactification radius

in general, still unclear what info is encoded and whether it can uniquely specify the topological order at hand

situation not much better for non-interacting systems--- even here, unclear what information is encoded and how one can read it off from the spectrum

here, we show that the information about the Berry phase can be completely recovered from the entanglement spectrum and provide a simple formula to do so

This can be related to earlier attempts, REFS

\section{Model}

@@

\section{Geometric phase and Entanglement spectrum}
Consider a quadratic Hamiltonian in one dimension
\begin{equation}
\mathcal{H} = \sum_{ij,\alpha\beta} c_{i\alpha}^\dagger H_{ij,\alpha \beta}c_{j\beta}.
\end{equation}
We can obtain the single-particle eigenstates as
\begin{align*}
\sum_{j\beta}H_{ij,\alpha\beta} u^{p\mu}_{j\beta} = E_{p\mu} u_{i\alpha}^{p\mu},
\end{align*}
where $[U]_{j\beta,p\mu} = u^{p\mu}_{j\beta}$ is the unitary matrix that diagonalizes $H$. The polarization can be obtained as the many-body average of position operator, regularized for periodic boudnary conditions \cite{Resta1997}
\begin{align*}
\expval{X} = -\frac{L}{2\pi} \rm{ Im\, Ln }\bra{\Psi_0}e^{i \frac{2\pi}{L}\hat{X}}\ket{\Psi_0},
\end{align*}
where $\ket{\Psi_0}$ is the many-body ground state and $\hat{X}$ is the sum of all the single-particle position operators.
Using that the ground state is a Slater determinant of the single-particle eigenstates, $\ket{u^{p\mu}}$, it can be rewritten as
\begin{align*}
\expval{\hat{X}} = -\frac{L}{2\pi} \rm{Im \, Ln}\, {\rm det} \, \tilde{S},
\end{align*}
where the matrix $\tilde{S}$ is obtained from another matrix S with elements
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha}u^{p\mu\, \ast}_{j \alpha} e^{-i\frac{2\pi}{L}j}u^{q\nu}_{j \alpha}
\end{equation}
by restricting it to the space of occupied single-particle states. The geometric phase can now be obtained as
\begin{equation}
\gamma = \expval{\hat{X}}\frac{2\pi}{L}
\end{equation}

The correlation function in position space is given by
\begin{align*}
C_{ij}^{\alpha \beta} = \expval{c_{i\alpha}^\dagger c_{j\beta}}.
\end{align*}
In terms of the fermions that diagonalize the Hamiltonian, 
\begin{align*}
& \gamma_{i\alpha} = \sum_{j\beta}u_{j\beta}^{i\alpha \, \ast} c_{j\beta} \\
& c_{i\alpha} = \sum_{j\beta} u^{j\beta}_{i\alpha} \gamma_{j\beta}
\end{align*}
we have
\begin{align*}
C_{ij}^{\alpha \beta} =& \sum_{pq,\mu\nu} u^{q\nu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{q\nu} } \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{p\mu} } \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} [1-{\rm sign}(E_{p\mu})]/2.
\end{align*}
The first term is simply a kronecker delta between both sets of indeces. The second term can be rewritten as
\begin{align*}
UD(\abs{D})^{-1}U^\dagger =& UD(D^2)^{-1/2}U^\dagger \\
=& U D U^\dagger (U D^2 U^\dagger)^{-1/2} \\
=& H(H^2)^{-1/2},
\end{align*}
where $D$ is the diagonal Hamiltonian, $H=UDU^\dagger$. Finally the correlation function can be obtained in position space as
\begin{align*}
C_{ij}^{\alpha \beta} =& \frac{1}{2}\left[I - H/ (H^2)^{-1/2} \right]_{ij, \alpha \beta}
\end{align*}
The entanglement spectrum (ES) is then obtained as the spectrum of the subsystem correlation function obtained after bisecting the system into two equal parts.

@@

For symmetry-protected topological systems in one dimension there is a well-known relation between ES and the geometric phase. The geometric phase, which is quantized, is zero whenever there is an even number per edge of eigenvalues at $\xi = 1/2$, and $\pi$ when the number is odd. Beyond symmetry protected systems there have been observations of similarities between the midgap eigenvalue of the entanglement spectrum and the geometric phase \cite{Ryu2006,Huang2012,Huang2012-2}, which in the case of the fully dimerized model becomes an exact identity \cite{Ryu2006}.

Consider the combination of all eigenvalues localized in one of the edges modulo one,
\begin{equation}
\chi = \sum_{i \in l} \xi_l \quad {\rm mod} \quad 1. 
\end{equation}
As we will show below, for systems where the ES has more than one edge state per edge one can find the geometric phase in the thermodynamic limit as
\begin{equation}
\lim_{L \rightarrow \infty}\gamma/2\pi =\lim_{L \rightarrow \infty} \chi
\label{mainl}
\end{equation}
In the examples studied in references \cite{Ryu2006,Huang2012,Huang2012-2} the main contribution comes from one midgap state, while the rest of the edge states sit close to the bulk bands and therefore give a small contribution, which explains why the relation between the midgap eigenvalue and the geometric phase was almost an identity and why it breaks as we go deep into the trivial state. Alternatively, using that the entanglement spectrum is particle-hole symmetric for homogeneous systems we can express $\chi$ as
\begin{equation*}
\chi = \frac{1}{2} \left( \sum_{i\in l}\xi_i-\sum_{i\in r}\xi_i \pm \sum_{i\in b}\xi_i  + N \right) \quad {\rm mod} \quad 1,
\end{equation*}
where $N$ is the total number of electrons in the ground state and the sign on the bulk term is irrelevant since $-1/2 \, {\rm mod} \, 1 = 1/2 \, {\rm mod} \, 1$. This form is preferable when doing numerics because it is difficult to differentiate between edge modes and bulk modes, since the latter can acquire a finite eigenvalue due to finite size effects or numerical errors. In Fig.\ref{bdi20} we show for several cases how Eq. \ref{mainl} holds up to finite size errors. $\chi$ is equivalent to the trace of the ES of a semi-infinite chain, but one can extract the information from the ES of a ring in the thermodynamic limit by looking at the localization structure of the eigenstates, only looking at the spectrum is not enough. One drastic example of this can be seen in figures \ref{bdi20}.b and \ref{bdi20}.c , where two ES that look almost identical result in a very different geometric phase.

@@

\makebox[0pt]{
\subfloat[$t = 0, t' = 2, \kappa = 10^{-5}$]{
  \includegraphics[width=35mm]{2_c.pdf}
}
\subfloat[$t = 0, t' = 2, \kappa = 0.3$]{
  \includegraphics[width=35mm]{2_d.pdf}
}
}
\caption{fill. }
\label{bdi20}
\end{figure}
as numerically is
This relation between the geometric phase and the entanglement spectrum is a consequence of an identity between the geometric phase and the many-body entanglement spectrum previously found for infinite chains \cite{Zaletel2014}. In Appendix A we modified this derivation to account for the periodic boundary conditions we use and show how their result reduces to Eq. \ref{mainl} when expressing it in terms of the ES.

@@

In reference \cite{Zaletel2014} translational invariance was used to derive Eq. \ref{zaletel} in Appendix A. Say we break translational invariance by adding weak disorder to the system. This will modify slightly the entanglement spectrum but not the charge assignment. Without translational invariance the entanglement spectrum will depend on where is the cut placed on the ring, even though the ground state, and therefore the geometric phase, will not. One thing one could do to get rid of the cut-dependence is to perform an average over all the spatial cuts of expression \ref{mainl}. Defining now the cut-dependent $\chi$ as 
\begin{equation*}
\chi_s = \frac{1}{2} \left( \sum_{i\in l}\xi^s_i-\sum_{i\in r}\xi^s_i \pm \sum_{i\in b}\xi^s_i  + N \right) \quad,
\end{equation*}
for a cut in positions $s$ and $s+L/2$. We can write
\begin{equation}
\gamma = \expval{\chi_s \, {\rm mod}\, 1}_s \, {\rm mod}\, 1.
\label{modmod}
\end{equation}
The average of a circular quantity is however not well-defined. For the case when the width of $\chi_s$ is lower than one, there is a range of values $[x,x+1]$ where $\chi_s$ does not cross $0,1$, and averaging over this range makes sure than once we take the second modulo $1$ all the different possible results for the cut-average are reduced to a single one. The shortcomings of expression \ref{modmod} will be discussed later, but to show that this cut-average is correct we will consider now a system in the topological phase to which we add weak disorder, so one can then take the cut-average around $[0,1]$. In Fig.\ref{disorder} we show $1 - \expval{\chi_s \, {\rm mod}\, 1}_s \, {\rm mod}\, 1 /\gamma$ in red for different system sizes. In blue we have shown the same difference using a disorder average. Both values match fairly well and they seem to vanish in the thermodynamic limit, scaling as $O(L^{-2})$. The fact that different disorder configurations match so well points to a systematic finite-size error in the way we calculate either the ES or the geometric phase that is system-independent. From this result one could think that the cut-average is some sort of disorder average. The implication however goes in the opposite direction. If the cut-average is assumed to be correct, that implies that in the thermodynamic limit we can group all the infinite different disorder configurations into an infinite set of $L$ configurations relating to different cuts of the same system, which average individually to the correct $\gamma_{\rm conf}$.

For stronger disorder one cannot be sure if $\chi_s$ crosses $0,1$ or not so this method cannot be applied but for continuous inhomogeneous perturbations one can follow $\chi_s$, so we can use expression \ref{modmod} even when they are strong. In Fig.\ref{4} we show the results using a position dependent $\kappa$ parameter, shown in figure \ref{4}.a for different values of $m$ and the amplitude of $\kappa$. In cases Fig.\ref{4}.b,c the width of $\chi_s$ is small so we can average taking the value ranges $[0,1]$ and $[-1/2,1/2]$ respectively and we see that the cut average matches the geometric phase perfectly. However for stronger perturbations, as shown in Fig.\ref{4}.d, one can have that the width of $\chi_s$ is bigger than 1 and there is no meaningful range of values one can average. Instead what one can do is to recover $\chi_s$ (\carlos{ only edge modes}) by constructing a new object $\tilde{\chi}_s$ by adding or substracting $1$ each time $\chi_s$ crosses $0,1$, taking the cut-average of $\tilde{\chi}_s$ and then take the modulo 1.

@@

\caption{ fill. put .5 line, also break the symmetry. with an envelope.}
\label{4}
\end{figure}
(We use a profile which is invariant for $x\rightarrow x+L/2$ for practical reasons, so we can illustrate all cases, as it enhances the effect of the nonhomogeneity. But this symmetry is not necesarry for anything we discuss here)
The last case with an easy solution is when the system is fully gapped, so that the ES doesn't have modes crossing $1/2$ for any cut. In this case one can make use of the particle-hole symmetry of the ES and use only the lower half of the spectrum to build $\chi_s$ with no issues due to the bulk contribution. 
One can explot the fact that, even in the inhomogeneous case, the ES has a type of particle-hole symmetry. If $\xi$ is an eigenvalue for a cut in $s$ then $1-\xi$ is an eigenvalue of the cut in $s+L/2$. This means that all the information is already contained in half of the spectrum. Using the lower half of the spectrum ensures that we have all the information about the edge eigenstates but the bulk contribution vanishes. This is however only valid when there is a gap between the two halfs of the spectrum at any cut $s$. MERGE
Another interesting point is that the two charge choices now leave to different $\chi_s$ since the particle-hole is no longer fulfilled for individual cuts. However both still lead to the same result because there is still a symmetry involving cuts related by a half-system translation. Interestingly for $\pm 1/2$ the quantity $\chi_s$ seems to be smaller, leading to fewer issues when averaging.

>>>>

\section{Introduction}
Topological phases of matter have attracted more and more attention during the last decades, not the least because many more relevant systems have been realized experimentally. 
The early focus was mainly on \emph{topologically ordered} systems \cite{wenbook}, where interaction effects are crucial for stabilizing the phase. 
The most notable examples are the fractional quantum Hall effect~\cite{Tsui1982} and quantum spin liquids~\cite{Balents2010spin}. 
However, since 2005~\cite{kane2005quantum,roy2009topological} the focus has shifted to non-interacting topological phases, which can be characterized in terms of topological invariants~\cite{ryu2010topological}. 
Symmetries are often necessary to protect these topological phases, and determine which distinct topological phases can be realized for a given dimensionality.

There are a few tools to identify topological phases. 
A very efficient one is the entanglement entropy~\cite{}, which allows one to identify the total quantum dimension of the underlying topological field theory. 
As such, the entanglement entropy can only be used for interacting systems. 
Its main drawback is the fact that it can only provide one of the quantum numbers, and is not able to uniquely determine the topologically ordered phase at hand. 
In addition, it requires a scaling analysis, which can be challenging for strongly correlated systems. 
Another, closely related tool, is the entanglement spectrum (ES), originally introduced for fractional quantum Hall systems~\cite{Li2008entanglement}. 
It was conjectured to provide information about the edge spectrum, which was later shown to be a general feature of ground states with an effective topological field theory description~\cite{Qi2012general}. 
It also proved to be an effective tool for fractional Chern insulators~\cite{Regnault2011fractional} and certain quantum spin liquids~\cite{yao2010entanglement}. \mariac{Others?}

Generically, any of the proposed tools provides some information about the phase naturally, while others seem inaccessible. 
It is not known, whether there exists a method to obtain the full topological data from the ground state. 
For the simplest quantum Hall states, the Laughlin states,  it was shown that the ES does, in fact, contain the full information about the topological phase~\cite{hermanns2011haldane}.
However, it is far from clear if this holds for more complicated/nonabelian states as well.

Here, we address a much simpler question: what information can be extracted from the ES of \emph{non-interacting} systems?
The latter can be computed very efficiently, using methods developed by Peschel and others~\cite{Peschel2003}. 
It was subsequently shown for topological insulators/superconductors that the ES for (gapped) periodic systems is equivalent to the flat-band energy spectrum of the corresponding system with open boundaries~\cite{Fidkowski2010entanglement}. 
The same correspondence was also found for closely related gapless systems~\cite{matern2018entanglement}
However, even for non-interacting systems, it is unclear which information (beyond the  `edge' spectrum) is encoded in the spectrum.

In this manuscript, we show that the information about the Berry phase can be completely recovered from the entanglement spectrum and provide a simple formula to do so. 
We relate our results to earlier ones XXX

\emph{Outline of the paper}
\section{Model.}

@@

\section{Geometric phase and Entanglement spectrum}
Consider a quadratic Hamiltonian in one dimension
\begin{equation}
\mathcal{H} = \sum_{ij,\alpha\beta} c_{i\alpha}^\dagger H_{ij,\alpha \beta}c_{j\beta}.
\end{equation}
We can obtain the single-particle eigenstates as
\begin{align*}
\sum_{j\beta}H_{ij,\alpha\beta} u^{p\mu}_{j\beta} = E_{p\mu} u_{i\alpha}^{p\mu},
\end{align*}
where $[U]_{j\beta,p\mu} = u^{p\mu}_{j\beta}$ is the unitary matrix that diagonalizes $H$. The polarization can be obtained as the many-body average of position operator, regularized for periodic boudnary conditions \cite{Resta1997}
\begin{align*}
\expval{X} = -\frac{L}{2\pi} \rm{ Im\, Ln }\bra{\Psi_0}e^{-i \frac{2\pi}{L}\hat{X}}\ket{\Psi_0},
\end{align*}
where $\ket{\Psi_0}$ is the many-body ground state and $\hat{X}$ is the sum of all the single-particle position operators.
Using that the ground state is a Slater determinant of the single-particle eigenstates, $\ket{u^{p\mu}}$, it can be rewritten as
\begin{align*}
\expval{\hat{X}} = -\frac{L}{2\pi} \rm{Im \, Ln}\, {\rm det} \, \tilde{S},
\end{align*}
where the matrix $\tilde{S}$ is obtained from another matrix S with elements
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha}u^{p\mu\, \ast}_{j \alpha} e^{-i\frac{2\pi}{L}j}u^{q\nu}_{j \alpha}
\end{equation}
by restricting it to the space of occupied single-particle states. The geometric phase can now be obtained as
\begin{equation}
\gamma = \expval{\hat{X}}\frac{2\pi}{L} + \frac{1}{2}
\end{equation}
\carlos{There is an extra 1/2 factor compared to Kivelson's formula which I don't know where it is coming from.}
The correlation function in position space is given by
\begin{align*}
C_{ij}^{\alpha \beta} = \expval{c_{i\alpha}^\dagger c_{j\beta}}.
\end{align*}
In terms of the fermions that diagonalize the Hamiltonian, 
\begin{align*}
& \gamma_{i\alpha} = \sum_{j\beta}u_{j\beta}^{i\alpha \, \ast} c_{j\beta} \\
& c_{i\alpha} = \sum_{j\beta} u^{j\beta}_{i\alpha} \gamma_{j\beta}
\end{align*}
we have
\begin{align*}
C_{ij}^{\alpha \beta} =& \sum_{pq,\mu\nu} u^{q\nu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{q\nu} } \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{p\mu} } \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} [1-{\rm sign}(E_{p\mu})]/2.
\end{align*}
The first term is simply a kronecker delta between both sets of indeces. The second term can be rewritten as
\begin{align*}
UD(\abs{D})^{-1}U^\dagger =& UD(D^2)^{-1/2}U^\dagger \\
=& U D U^\dagger (U D^2 U^\dagger)^{-1/2} \\
=& H(H^2)^{-1/2},
\end{align*}
where $D$ is the diagonal Hamiltonian, $H=UDU^\dagger$. Finally the correlation function can be obtained in position space as
\begin{align*}
C_{ij}^{\alpha \beta} =& \frac{1}{2}\left[I - H/ (H^2)^{-1/2} \right]_{ij, \alpha \beta}
\end{align*}
The entanglement spectrum (ES) is then obtained as the spectrum of the subsystem correlation function obtained after bisecting the system into two equal parts.

@@

For symmetry-protected topological systems in one dimension there is a well-known relation between ES and the geometric phase. The geometric phase, which is quantized, is zero whenever there is an even number per edge of eigenvalues at $\xi = 1/2$, and $\pi$ when the number is odd. Beyond symmetry protected systems there have been observations, in simple systems with one edge-mode per edge, of similarities between the midgap eigenvalue of the entanglement spectrum and the geometric phase \cite{Ryu2006,Huang2012,Huang2012-2}, which in the case of the fully dimerized model becomes an exact identity \cite{Ryu2006}.

Consider the combination of all eigenvalues localized in one of the edges modulo one,
\begin{equation}
\chi = \sum_{i \in r} \xi_i \quad {\rm mod} \quad 1. 
\end{equation}
As we will show below, one can find the geometric phase in the thermodynamic limit as
\begin{equation}
\lim_{L \rightarrow \infty}\gamma/2\pi =\lim_{L \rightarrow \infty} \chi.
\label{mainl}
\end{equation}
In the examples studied in references \cite{Ryu2006,Huang2012,Huang2012-2} the main contribution comes from one midgap state, while the rest of the edge states sit close to the bulk bands and therefore give a small contribution, which explains why the relation between the midgap eigenvalue and the geometric phase was almost an identity and why it breaks as we go deep into the trivial state. Alternatively, using that the entanglement spectrum is particle-hole symmetric for homogeneous systems we can express $\chi$ as
\begin{equation*}
\chi = \frac{1}{2} \left( \sum_{i\in r}\xi_i-\sum_{i\in l}\xi_i \pm \sum_{i\in b}\xi_i  + N_A \right) \quad {\rm mod} \quad 1,
\end{equation*}
where $N_A$ is the total number of electrons in subsystem A  in the ground state and the sign on the bulk term is irrelevant since $-1/2 \, {\rm mod} \, 1 = 1/2 \, {\rm mod} \, 1$. This form is preferable when doing numerics because due to finite size effects the bulk modes can acquire a finite eigenvalue and therefore in general it is difficult to classify between bulk and edge modes. \carlos{There is also the issue edge-modes and bulk-modes mixing in the inhomogenous case, introduce later?} In practice we diagonalize $C_A + \lambda C_A\hat{X}C_A$, where $\hat{X}$ is the position operator and $\lambda$ is a small parameter. The second term ensures that the originally degerate bulk states are also localized, so we can divide all eigenvalues in right and left sectors and assign the charges $1/2$ and $-1/2$, respectively. In Fig.\ref{bdi20} we show for several cases how Eq. \ref{mainl} holds up to finite size errors. $\chi$ is equivalent to the trace of the ES of a semi-infinite chain, but one can extract the information from the ES of a ring in the thermodynamic limit by looking at the localization structure of the eigenstates, only looking at the spectrum is not enough.

One drastic example of this can be seen in figure \ref{old} , where two ES that look almost identical result in a very different geometric phase.

@@

\makebox[0pt]{
\subfloat[$t = 0, t' = 2, \kappa = 10^{-5}$]{
  \includegraphics[width=35mm]{2_c.pdf}
}
\subfloat[$t = 0, t' = 2, \kappa = 0.3$]{
  \includegraphics[width=35mm]{2_d.pdf}
}
}
\caption{fill. }
\label{bdi20}
\end{figure}
\begin{figure}[h!]
\centering
\makebox[0pt]{
\subfloat[$t = 0, t' = 2, \kappa = 10^{-5}$]{
  \includegraphics[width=35mm]{a20.pdf}
}
\subfloat[$t = 0, t' = 2, \kappa = 0.3$]{
  \includegraphics[width=35mm]{d20.pdf}
}
}
\caption{Old plot, do I keep it? }
\label{old}
\end{figure}

This relation between the geometric phase and the entanglement spectrum is a consequence of an identity between the geometric phase and the many-body entanglement spectrum previously found for infinite chains \cite{Zaletel2014}. In Appendix A (\carlos{I have to take a look at it again}) we modified this derivation to account for the periodic boundary conditions we use and show how their result reduces to Eq. \ref{mainl} when expressing it in terms of the ES.

@@

In reference \cite{Zaletel2014} translational invariance was used to derive Eq. \ref{zaletel} in Appendix A. Say we break translational invariance by adding weak disorder to the system. This will modify slightly the entanglement spectrum but not the charge assignment. Without translational invariance the entanglement spectrum will depend on where is the cut placed on the ring, even though the ground state, and therefore the geometric phase, will not. One thing one could do to get rid of the cut-dependence is to perform an average over all the spatial cuts of expression \ref{mainl}. Defining now the cut-dependent $\chi$ as 
\begin{equation*}
\chi_s = \frac{1}{2} \left( \sum_{i\in r}\xi^s_i-\sum_{i\in l}\xi^s_i \pm \sum_{i\in b}\xi^s_i  + N_A \right) \quad,
\end{equation*}
for a cut in positions $s$ and $s+L/2$. We can write
\begin{equation}
\gamma = \expval{\chi_s}_{\rm cuts} \, {\rm mod}\, 1.
\label{gphase_ih_mod}
\end{equation}
However this expression is not practical. In order for the bulk contribution to vanish with the modulo $1$ it has to be constant for different cuts. Last section we divided the bulk modes in right and left, but now we encounter the problem that their localization changes with the cut, so that there is the possibility that one bulk mode around $1/2$ changes its charge, leading to a wrong result. To get around this problem we can get rid of the bulk contribution before averaging,
\begin{equation}
\gamma = \expval{\chi_s \, {\rm mod}\, 1}_s \, {\rm mod}\, 1,
\label{gphase_ih_modmod}
\end{equation}
however the average of a circular quantity is not well defined. We will explain how one can perform the cut-average for different cases so that one can recover expression \ref{gphase_ih_mod} from expression \ref{gphase_ih_modmod}.

For the case when the width of $\chi_s$ is lower than one, there is a range of values $[x,x+1]$ where $\chi_s$ does not cross $0,1$, and averaging over this range makes sure than once we take the second modulo $1$ all the different possible results for the cut-average are reduced to a single one. The shortcomings of expression \ref{modmod} will be discussed later, but to show that this cut-average is correct we will consider now a system in the topological phase to which we add weak disorder, so one can then take the cut-average around $[0,1]$. In Fig.\ref{disorder} we show $1 - \expval{\chi_s \, {\rm mod}\, 1}_s \, {\rm mod}\, 1 /\gamma$ in red for different system sizes. In blue we have shown the same difference using a disorder average. Both values match fairly well and they seem to vanish in the thermodynamic limit, scaling as $O(L^{-2})$. The fact that different disorder configurations match so well points to a systematic finite-size error in the way we calculate either the ES or the geometric phase that is system-independent. From this result one could think that the cut-average is some sort of disorder average. The implication however goes in the opposite direction. If the cut-average is assumed to be correct, that implies that in the thermodynamic limit we can group all the infinite different disorder configurations into an infinite set of $L$ configurations relating to different cuts of the same system, which average individually to the correct $\gamma_{\rm conf}$.

For stronger disorder one cannot be sure if $\chi_s$ crosses $0,1$ or not so this method cannot be applied but for continuous inhomogeneous perturbations one can follow $\chi_s$, so we can use expression \ref{modmod} even when the inhomogeneity is strong. In Fig.\ref{4} we show the results using a position dependent $\kappa$ parameter, shown in figure \ref{4}.a. In cases Fig.\ref{4}.b,c the width of $\chi_s$ is small so we can average taking the value ranges $[0,1]$ and $[-1/2,1/2]$ respectively and we see that the cut average matches the geometric phase perfectly. However for stronger perturbations, as shown in Fig.\ref{4}.d, one can have that the width of $\chi_s$ is bigger than 1 and there is no meaningful range of values one can average. Instead what one can do is to "double the periodicity of $\chi_s$" by adding or substracting $1$ each time it crosses $0,1$, taking the cut-average of and then take the modulo 1.

@@

\caption{ fill. put .5 line, also break the symmetry. with an envelope.}
\label{4}
\end{figure}

In Fig.\ref{4} we use a profile which is invariant for $x\rightarrow x+L/2$ because it enhances the effect of the nonhomogeneity, but this symmetry is not necesarry for anything we discuss here.

Another case where one can find a solution to the issue when cut-averaging is when the system is fully gapped, so that the ES doesn't have modes crossing $1/2$ for any cut. In this case one can make use of the particle-hole symmetry of the ES (If $\xi$ is an eigenvalue for a cut in $s$ then $1-\xi$ is an eigenvalue of the cut in $s+L/2$) and use only the lower half of the spectrum to build $\chi_s$ so that the bulk contribution vanishes enitrely. 
One can explot the fact that, even in the inhomogeneous case, the ES has a type of particle-hole symmetry.

Another interesting point to note is that the two charge choices now leave to different $\chi_s$ since the particle-hole symmetry is no longer fulfilled for individual cuts. However both still lead to the same result because of the particle-hole time $L/2$ translation symmetry. Interestingly for the half-integer charge choice the quantity $\chi_s$ seems to be smaller, leading to fewer issues when averaging.
